An Empirical Analysis of the Role of Energy in Economic Growth

### 2015-01-03 Articles/Article_EnerEcon_2014/EconGrowth3.Rnw
Added a note about data not determining coeffs precisely.

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \, \sigma_1 = \infty$), 
is a troubling characteristic of the estimated CES coefficients 
that is rarely investigated in the literature.
Indeed, for some economies (Iran and Zambia), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Note that this result is unlikely to be caused by poor-quality data alone. 
Presumably the best data in the world 
(United States, United Kingdom, and Japan) exhibits large uncertainty.

Some of the uncertainty is explained by the structure of the CES model itself. 
If $\delta$ or $\delta_1$ achieves an extreme value (0 or 1), 
the corresponding elasticity of substitution ($\sigma$ and $\sigma_1$) becomes meaningless. 
(See Equations~\ref{eq:CES}--\ref{eq:CESekl}.) 
Thus, large uncertainty in $\sigma$ and $\sigma_1$ is expected for any resampled model 
in which parameter estimates are positioned on the left or right boundaries 
of Figures~\ref{fig:CES_sigma1_delta1_graph} or~\ref{fig:CES_sigma_delta_graph}.

Some of the uncertainty in parameter estimates may be explained by structural change in the economies. 
For example, 
**** Martin: can you select one of the highly uncertain economies and discuss it? 
Perhaps Iran or Zambia? ****. 
To account for the large uncertainty, 
it may be useful to perform piecewise analyses of smaller time periods. 
For the present study, doing piecewise analysis of, say, 
Iran **** use the economy that Martin discusses above **** 
would mean that the CES model is estimated on two 10-year periods. 
Considering that there are six fitted parameters in the CES model 
($\gamma$, $\lambda$, $\delta$, $\delta_1$, $\rho$, and $\rho_1$), 
ten years of historical data leaves precious few 
(only 4) degrees of freedom for the fitting process.

@@

Based on the results and discussions above 
regarding the relative importance of and 
the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
should be undertaken with utmost caution. 
Indeed, rigorous statistical testing on the relative importance 
of production factors and elasticities of substitution is highly recommended. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies 
to enhance the reliability of results. 
**** Would econometrically specifying production function relationships and
performing sensitivity analysis really ``enhance the reliability'' of results? 
Or, would doing so improve transparency? --MKH ****

@@

The LINEX model is intriguing because of its thermodynamic roots, 
but our implementation suffers from 
poor reproduction of historical data and
difficulty in respecting parameter constraints.
If the LINEX function is to be used, 
we recommend that \citeapos{Kummel:2011aa} fitting approach 
(energy quantified by primary thermal energy and logistic curves for $a_0$ and $c_t$) 
or \citeapos{Ayres:2010ug} approach (energy quantified by useful work and constant values for $a_0$ and $c_t$) 
be employed 
and that the results be 
thoroughly tested and compared 
against other production functions.

>>>>

That so many estimates of elasticity of substitution exhibit large uncertainty, 
often spanning the range from perfect complements ($\sigma, \, \sigma_1 = 0$) 
to perfect substitutes ($\sigma, \, \sigma_1 = \infty$), 
is a troubling characteristic of the estimated CES coefficients 
that is rarely investigated in the literature.
Indeed, for some economies (Iran and Zambia), 
estimating any elasticity of substitution at all is problematic and fraught with peril. 
Earlier, we noted that 
even when residuals for the fit to historical data are small 
and the model produces fitted values that track the historical data closely, 
it may yet be difficult to estimate some or all of the model parameters precisely.
That appears to be the case for 
values of $\rho$ and $\rho_1$ and associated
elasticities of substitution~($\sigma$ and $\sigma_1$)
in the CES production function.
Note that this result is unlikely to be caused by poor-quality data alone. 
Elasticities of substitution for the economies with presumably the best data in the world 
(United States, United Kingdom, and Japan) exhibit large uncertainty.

Some of the uncertainty in parameter estimates is explained by 
the mathematical structure of the CES model itself. 
Equations~\ref{eq:CES}--\ref{eq:CESekl} indicate that 
if $\delta$ or $\delta_1$ achieves an extreme value (0 or 1), 
the corresponding elasticity of substitution ($\sigma$ or $\sigma_1$) becomes meaningless. 
If $\delta$ approaches unity, values of $\delta_1$ and $\sigma_1$ become meaningless.
Thus, large uncertainty in $\sigma$, $\sigma_1$, and $\delta_1$ can be expected for any CES model 
in which parameter estimates are positioned on the left or right boundaries
of Figures~\ref{fig:CES_sigma1_delta1_graph} or~\ref{fig:CES_sigma_delta_graph}.

Some of the uncertainty in parameter estimates may be explained by structural change in the economies. 
For example, 
**** Martin: can you select one of the highly uncertain economies and discuss it? 
Perhaps Iran or Zambia? ****. 
To account for the large uncertainty, 
it may be useful to perform piecewise analyses of smaller time periods. 
For the present study, doing piecewise analysis of, say, 
Iran **** use the economy that Martin discusses above **** 
would mean that the CES model is estimated on two 10-year periods. 
Considering that there are six fitted parameters in the energy-augmented CES model 
($\gamma$, $\lambda$, $\delta$, $\delta_1$, $\rho$, and $\rho_1$
in Equations~\ref{eq:CESkle}--\ref{eq:CESekl}), 
ten years of historical data leaves precious few 
(only 4) degrees of freedom for the fitting process.

@@

Based on the results and discussions above 
regarding (a)~the relative importance of factors of production and 
(b)~the elasticity of substitution among factors of production, 
we conclude that use of the CES model for energy analysis 
(Equations \ref{eq:CESkle}--\ref{eq:CESekl})
should be undertaken with utmost caution. 
Our results support \citeapos{Lecca:2011kr} conclusion 
for the need to econometrically specify production function relationships and 
to perform sensitivity analysis on nesting choices in economies 
to enhance the reliability of results. 
**** Would econometrically specifying production function relationships and
performing sensitivity analysis really ``enhance the reliability'' of results? 
Or, would doing so improve transparency? --MKH ****
Indeed, rigorous statistical testing on the relative importance 
of production factors and elasticities of substitution is strongly recommended.

@@

The LINEX model is intriguing because of its thermodynamic roots, 
but our implementation suffers from 
poor reproduction of historical data and
difficulty in respecting parameter constraints.
If the LINEX function is to be used for energy analysis, 
we recommend that either \citeapos{Kummel:2011aa} fitting approach 
(energy quantified by primary thermal energy and logistic curves for $a_0$ and $c_t$) 
or \citeapos{Ayres:2010ug} fitting approach 
(energy quantified by useful work and constant values for $a_0$ and $c_t$) 
be employed 
and that the results be 
thoroughly tested and compared 
against other production functions.
Doing so is left as future work.
